---
title: "Which Sectors Won? S&P 500 Sector Performance, 2025–2026"
author: "Ben Beauvais"
format: html
---

<!--
Sector analysis
-->

## Research question

How did the 11 S&P 500 sectors perform from January 2025 to today?
Which sectors gave the best return, and did the higher-return sectors also take on more risk?

**Data:** daily adjusted closing prices for the 11 Select Sector SPDR ETFs plus SPY (the whole S&P 500), pulled from Yahoo Finance with the `tidyquant` package.

## Setup

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(tidyquant)
```

## Get the data here: -->

```{r}
#| message: false
sectors <- tibble(
  symbol = c("XLK", "XLF", "XLE", "XLV", "XLY", "XLP",
             "XLI", "XLB", "XLU", "XLRE", "XLC", "SPY"),
  sector = c("Technology", "Financials", "Energy", "Health Care",
             "Consumer Disc.", "Consumer Staples", "Industrials",
             "Materials", "Utilities", "Real Estate",
             "Communication", "S&P 500 (SPY)")
)

prices <- tq_get(sectors$symbol, from = "2025-01-01") |>
  select(symbol, date, adjusted) |>
  left_join(sectors, by = "symbol")

head(prices)
```

## Wrangle: daily returns and growth of $100 to put the grwoth in perspective

```{r}
returns <- prices |>
  group_by(sector) |>
  arrange(date, .by_group = TRUE) |>
  mutate(
    daily_return = adjusted / lag(adjusted) - 1,   # % change from yesterday
    growth_100   = 100 * adjusted / first(adjusted) # value of $100 invested on day 1
  ) |>
  ungroup() |>
  filter(!is.na(daily_return))

summary_tbl <- returns |>
  group_by(sector) |>
  summarize(
    total_return = last(growth_100) / 100 - 1,
    volatility   = sd(daily_return) * sqrt(252)    # annualized
  ) |>
  arrange(desc(total_return))

summary_tbl
```

## 1. Univariate: what does a "normal day" look like for the S&P 500?

```{r}
returns |>
  filter(symbol == "SPY") |>
  ggplot(aes(x = daily_return)) +
  geom_histogram(binwidth = 0.0025, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Distribution of SPY daily returns since Jan 2025",
       x = "Daily return", y = "Number of days")
```

**My interpretation:** *(Spread is concentrated around 0 with some outliers around 12% and -4 - 7% which makes sense to me because the typical annual return for the s&p 500 is around 0?)*

## 2. Bivariate: which sectors are riskiest day to day?

```{r}
returns |>
  ggplot(aes(x = daily_return, y = fct_reorder(sector, daily_return, sd))) +
  geom_boxplot(fill = "lightgray", outlier.alpha = 0.3) +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Daily returns by sector (sorted by volatility)",
       x = "Daily return", y = NULL)
```

**My interpretation:** *(Higher volitility on stocks is met with higher returns, interesting to see the theory confirmed)*

## 3. Multivariate: growth of $100 over time

```{r}
returns |>
  ggplot(aes(x = date, y = growth_100, color = sector == "S&P 500 (SPY)")) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "gray50") +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ fct_reorder(sector, -growth_100, last)) +
  scale_color_manual(values = c("FALSE" = "gray30", "TRUE" = "firebrick"),
                     guide = "none") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b '%y") +
  labs(title = "Growth of $100 invested in each sector, Jan 2025 to today",
       subtitle = "Panels ordered from best to worst performer",
       x = NULL, y = "Value of $100")
```

**My interpretation:** *(Corresponds with the box plots with the higher the returns the more shaky the line)*

## 4. Risk vs. return: did more risk pay off?

```{r}
summary_tbl |>
  ggplot(aes(x = volatility, y = total_return, label = sector)) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_point(aes(color = sector == "S&P 500 (SPY)"), size = 3) +
  geom_text(size = 3, vjust = -0.9) +
  scale_color_manual(values = c("FALSE" = "gray30", "TRUE" = "firebrick"),
                     guide = "none") +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Risk vs. return by sector since Jan 2025",
       x = "Annualized volatility (risk)", y = "Total return")
```
